Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-06-22 write-up/covid-writeup.tex
Fixing theory

\subsection{Diagnostic versus serologic testing}
\label{section:testinginfo}

@@

\emph{Serological tests} look for an immunological response to the virus.  A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.  At this point, a serological test with perfect sensitivity will come back positive.  This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.

To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.  These studies are useful for disease surveillance. To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.  While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.

@@

Between April 25--29, 2020, Indiana conducted statewide random sample testing (both molecular and serological) of persons ages $\geq 12$ years to assess prevalence of active infection and presence of antibodies to SARS-CoV-2 (cite: Indiana report). A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. The same demographic data as provided by ISDH for Indian molecular testing data was included (e.g., summary statistics on age, sex, and race). Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.  In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

\subsubsection{Facebook/CMU symptom survey.}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users.  The survey is run by health researchers at Carnegie Mellon University. Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.

@@

We start with some simple notation.  Let $N$ denote the population size.  For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.  At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored.  We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

@@

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$.}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
$$
\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
As of June 10th, the US has performed $21,474,531$ tests.  The US population is roughly $312$ million, so $f = 0.069$.  The cumulative empirical prevalence is $9.3\%$ on May 8th. A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}. Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.  Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.  Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.  Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.  Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.  Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.
% \begin{wrapfigure}{r}{0.5\textwidth}

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions.
% First, we use publicly available auxiliary information to build selection propensities and consider an inverse probability weighting (IPW) estimator. We then extend the statistical error decomposition to the weighted setting. Second, we consider an extension of compartmental models from infectious disease epidemiology using death data.  After adjusting for measurement-error, this leads to an alternative, model-based forecast of prevalence. We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition.
The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This was proposed by~\cite{Chen2019} when covariates were observed on a random sample from the population.
% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
\begin{equation}
\label{eq:ipwest}
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
\end{equation}
where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

@@

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality. Third, the impact of measurement-error on data quality is changed when considering a weighted estimand. In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.

As demonstrated in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.  Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.   Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.

@@

In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
\begin{equation}
\label{eq:tvpropensity}
\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
\end{equation}
where $t=1,\ldots,T$ are the days when case-count data is reported.  Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.  For example, an individual who tests positive is unlikely to seek testing on subsequent days.  Moreover, an individual in a high prevalence area may be more likely to seek out testing.  Here, we assume that the covariate vector $X_{j,t}$ contains all features of past relevant for selection.

@@

\subsubsection{Uncertain sensitivity and specificity}

For prevalence there is some

For

For seroprevalence, \cite{Bendavid2020} report

, which leads to a range for $M$ of $(2.21, 2.38)$.  Therefore, the impact on bias of $\hat R_t$ appears moderate.  If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.  This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.

\subsubsection{Asymptotics}

@@

\begin{lemma}[Variance of IPW estimator] \normalfont
\label{lemma:ipw}
\textcolor{red}{Must check.}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

\subsection{Doubly robust estimation.}
Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error. In this context, the doubly-robust estimator is given by
$$
\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
$$
where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data. This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified. A statistical error decomposition can be derived
\begin{align}
\label{eq:statdecomp3}
\begin{split}
\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
\end{split}
\end{align}
See Appendix~\ref{app:DRderivation} for the derivation. Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.  First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components. Note that the model-based forecast does not impact the data quantity component.

If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size. Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.  For decent models, one would expect this ratio to be greater than one.  Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.  Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.

Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.  Here, we again use the probability samples and consider the doubly-robust estimator
$$
\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
$$
where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance. See Appendix~\ref{app:asympderivations} for additional details.

>>>>

\subsection{Diagnostic testing}
\label{section:testinginfo}

@@

% \emph{Serological tests} look for an immunological response to the virus.  A week or so after an individual is infected with SARS-CoV-2, the individual will start producing antibodies.  At this point, a serological test with perfect sensitivity will come back positive.  This test provides evidence of past infection while the molecular test provides evidence of an \emph{active} infection.

% To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.  These studies are useful for disease surveillance. To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.  While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.

@@

Due to testing restrictions and other potential selection biases, publicly reported covid testing data may not be sufficient to understand active infection rates or the disease trajectory.  Here, we discuss two random samples that can be leveraged to get either unbiased estimates of the statewide active infection rate or as auxiliary information to address selection bias in the testing data.

\subsubsection{Random statewide testing}

Between April 25--29, 2020, Indiana conducted statewide random molecular testing of persons ages $\geq 12$ years to assess prevalence of active infection to SARS-CoV-2~\citep{Yiannoutsos2021}. A stratified random sampling design was conducted using Indiana’s 10 public health preparedness districts as sampling strata. 15,495 participants were contacted resulting in a final sample size of 3,658. The same demographic data as provided by ISDH for Indian molecular testing data was included (e.g., summary statistics on age, sex, and race). Participants in the probability sample were also asked if they experienced any COVID-19 compatible symptoms during the past 2 weeks or had shared a household with someone who had a positive test result for SARS-CoV-2. During May 2--3, 2020, an additional nonrandom sample of 898 individuals was also collected.
% In this paper, the nonrandom sample is used to assess whether the cumulative summary statistics are appropriate in constructing the inverse-probability weights as well as in constructing appropriate weights using symptom information.

\subsubsection{Facebook/CMU symptom survey.}
Since April 2020, Facebook has started randomly showing an opt-in symptom survey to its users~\citep{delphisurvey}.  Data collected includes basic demographic information and if the respondent has symptoms such as fever, coughing, shortness of breath, or loss of smell which are associated with COVID-19.  The survey defines an individual as displaying \emph{COVID-like symptoms} if they exhibit a fever along with a cough, or shortness of breath, or difficulty breathing.  Figure~\ref{fig:fbsymptoms} displays nonparametric regression estimates of the fraction of individuals within age and gender strata.  In this paper, we use this as the main source of auxiliary information on time-varying characteristics of the population of Indiana, e.g., fraction of popopulation displaying symptoms or in contact with COVID-19 positive individuals).

@@

We start with some simple notation.  Let $N$ denote the population size.  At a given time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, the dynamic nature of the outbreak and recoverability of individuals are initially ignored.  We assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was tested ($I = 1$) or not ($I=0$).

To start, we assume the overall number of COVID-19 cases and/or disease prevalence are of primary interest. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem\footnote{equal probability of selection method} design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

@@

Tests are imperfect.  As discussed in Section~\ref{section:testinginfo}, COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  In discussions of inaccurate testing, the standard assumption is measurement error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and resulting errors can become magnified, muted, or even switch signs.

@@

\begin{figure}[!th]
\centering
\includegraphics[width = 0.4\textwidth]{../figs/mem_heatmap_article.png}
\caption{Imperfect testing adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.005$ and $FN=0.172$. Color scaled so blue = $-10$, white = $0$, and red = $10$. \textcolor{red}{NEED TO DO FOR FP = 0.05 and FN = 0.3}}
\label{fig:heatmap}
\vspace{-0.3cm}
\end{figure}

@@

A key question is ``What is the (effective) sample size from a SRS with perfect testing that would yield equivalent MSE to the current testing strategy?'' In Appendix~\ref{app:effss}, we show the effective sample size $n_{eff}$ can be bounded by
$$
\frac{f}{1-f} \times \frac{1}{E_{\I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
As of June 10th 2020, the US has performed $21,474,531$ tests.  The US population is roughly $312$ million, so $f = 0.069$.  \textcolor{red}{FIX: The cumulative empirical prevalence is $9.3\%$ on May 8th. A recent study estimates a false positive and negative rate for antibody testing at $0.5$\% and $17.2$\% respectively \citep{Bendavid2020}. Supposing COVID-19 positive individuals are $2$ times more likely to get tested, then the effective sample size is $92$.Recent proposals~\citep{Siddarth2020} have argued for increased testing capacity, which may likely reduce the relative sampling rate.  If the relative sampling rate drops to $1.2$ then the effective sample size will increase to $1950$.  Thus the effective sample size even in optimistic scenarios is equivalent to a small random sample from the population.  Moreover, increased testing capacity may also come at the cost of increases in false positive and negative rates.  Consider the case where $f$ increases from $0.05$ to $0.1$ and the increase is associated with the false negative rate rising from $5\%$ to $20$\%.  Then the effective sample size increases by a factor of $1.76$ rather than the expected $4$ fold increase.}
% \begin{wrapfigure}{r}{0.5\textwidth}

@@

The prior section presented negative consequences of selection-bias and measurement error when estimating infection prevalence from observed case-count data.  This raises the immediate question whether prevalence estimators can be improved.  In this section, we consider two directions.
The proposed methods and statistical error decompositions guide our recommendations in Section~\ref{section:discussion}.

@@

Here we assume access to a probability sample measuring both the outcome and exact same set of covariates.  Let $\tilde I_j$ be an indicator that the individual was included in the probability sample and $\tilde W_j$ be the probability of inclusion for the $j$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
\begin{equation}
\label{eq:auxinfoprob}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
\end{equation}
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  This extends the work by~\cite{Chen2019} which considered auxiliary information coming only from a simple random sample.
% Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

@@

Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
\begin{equation}
\label{eq:ipwest}
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
\end{equation}
where equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$.
% Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

@@

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Second, the relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality. Third, the impact of measurement-error on data quality is changed when considering a weighted estimand. In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.

As demonstrated below in Lemma~\ref{lemma:ipw}, if the propensity model is correctly specified then the $E_{\bf R} [ \rho_{\tilde I (X), Y} ] = 0$ and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the weights are not correctly specified then the data quality index is unlikely to inversely scale with population size.  Indeed, Horvitz-Thompson estimators are sensitive to errors in the estimated weights.  Low selection propensity imply large weights which can have deleterious impacts by certain observations dominating estimation.   Similar to~\cite{Meng2018}, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from an IPW estimator are likely to put too much confidence in the sheer size of the data.

@@

In prior sections on selection propensity estimation, we ignored the dynamic nature of the outbreak.  Here we extend the IPW approach to the temporal setting by considering the joint likelihood
\begin{equation}
\label{eq:tvpropensity}
\sum_{t=1}^T \left[ \sum_{j=1}^N I_{j,t} \log \left( \frac{\pi_t (X_{j,t}; \theta)}{1-\pi_t(X_{j,t}; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi_t (X_{j,t}; \theta) \right) \right]
\end{equation}
where $t=1,\ldots,T$ are the days when case-count data is reported.  Here, $I_{j,t}$ denotes self-selection into testing on day $t$, which is highly correlated with prior testing and results.  For example, an individual who tests positive is unlikely to seek testing on subsequent days.  Moreover, an individual in a high prevalence area may be more likely to seek out testing.  Here, we assume that the covariate vector $X_{j,t}$ contains all features of the past relevant for selection.

@@

%\subsubsection{Uncertain sensitivity and specificity}
%
%For prevalence there is some
%
%For
%
%For seroprevalence, \cite{Bendavid2020} report
%
%
%
%, which leads to a range for $M$ of $(2.21, 2.38)$.  Therefore, the impact on bias of $\hat R_t$ appears moderate.  If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.  This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
%
%
\subsubsection{Asymptotics}

@@

\begin{lemma}[Variance of IPW estimator] \normalfont
\label{lemma:ipw}
\textcolor{red}{Must check.}
The estimates~$\hat \mu_t := \bar y_{n,t}^\star$, $\hat \pi_{j,t}$, $\hat FP$, and $\hat FN$ are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - FP - (1-FP-FN) \cdot \mu_t}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

%\subsection{Doubly robust estimation.}
%Rather than relying solely on IPW or epidemiological forecasts, here we combine forecasting and inverse-probability weighting by extending recent work by~\cite{Chen2019} to account time-varying propensities and measurement-error. In this context, the doubly-robust estimator is given by
%$$
%\bar y_{n}^{(DR)} = \frac{1}{N} \sum_{j=1}^N \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
%$$
%where the estimates $\hat \mu(x_j)$ are not corrected for measurement-error as they are built to estimate true case count from death-only data. This estimator is called ``doubly-robust'' because it is consistent if either the model-based forecasts or the time-varying propensities are correctly specified. A statistical error decomposition can be derived
%\begin{align}
%\label{eq:statdecomp3}
%\begin{split}
%\rho_{\tilde I (X), Y-\mu(X)} &\times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y-\mu(X)}  \\
%&\times \underbrace{\left[ 1 - \frac{\rho_{\tilde I(X),Y} \sigma_{Y}}{\rho_{\tilde I(X),Y-\mu(X)} \sigma_{Y-\mu(X)}} \times \tilde \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{\tilde f_0 (1-\bar Y) + \tilde f_1 \bar Y} \right]}_{\tilde D_M}.
%\end{split}
%\end{align}
%See Appendix~\ref{app:DRderivation} for the derivation. Comparing~\eqref{eq:statdecomp3} to~\eqref{eq:statdecomp2} shows that the addition of the model-based focast impacts the estimation error in two ways.  First, if the model is adequate, then one may expect a reduction in the problem difficulty and (potentially) in the data quality components. Note that the model-based forecast does not impact the data quantity component.
%
%If the model-based forecast is correct then $E_{\bf R} [ \rho_{\tilde I (X), Y-\mu(X)} ] = 0$  and therefore $\rho_{\tilde I(X), Y} = O(N^{-1})$; however, if the forecasts are not correctly specified then the data quality index is unlikely to inversely scale with population size. Interestingly, the impact of measurement-error on data quality now depends on a relative comparison of the data quality and problem difficulty of the weighted estimator and the doubly-robust estimator.  For decent models, one would expect this ratio to be greater than one.  Therefore, $\tilde \Delta < 0$ implies an expected increase in impact of measurement-error and $\tilde \Delta > 0$ implies the magnitude could shrink toward zero or change sign and have an even large magnitude in the opposite direction.  Again, if the data quality is not at the level of $N^{-1}$, then confidence intervals constructed from a DR estimator are likely to put too much confidence in the sheer size of the data.
%
%Unfortunately, the first term cannot be computed as covariate information is not collected on the entire population.  Here, we again use the probability samples and consider the doubly-robust estimator
%$$
%\frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
%$$
%where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance. See Appendix~\ref{app:asympderivations} for additional details.
%
%\begin{lemma}[Variance of doubly-robust estimator] \normalfont
%\label{lemma:dr}
%\textcolor{red}{Must fix.}
%The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
%are solutions to the following set of estimating equations
%$$
%\Phi_n (\eta_t) =
%\left (
%\begin{array}{c}
%\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
%\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
%\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
